---
title: "ROC analysis of T vs. ME thresholds"
author: "Stacy Hung"
output:
  pdf_document: default
  html_notebook: default
---

## ############################################################
## Optimize variant calling thresholds for low tumor content samples
## ############################################################

Assess sensitivity and specificity of variant detection using tumor vs. constitutional as the gold standard.

Relevant definitions:
TP = variant is detected and part of the gold std dataset
TN = variant is not detected and not part of the gold std dataset
FP = variant is detected and not part of the gold std dataset
FN = variant is not detected, but part of the gold std dataset

Thresholds to assess (each as many different cutoffs as logical):
1. Tumor variant alelle frequency
2. Number of tumor variant reads

```{r}
library(plyr) # for revalue function
library(dplyr)
library(hashmap) # for hashmap function
library(ggplot2)
```

## Read in data and equalize theem for comparison (samples, genes, effects, etc.)

```{r}
# read in gold standard calls (i.e. from the exome) and map ids
gold.data <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/snvs_indels.full_cohort_minus_GZ229.default_and_optimized.FINAL_DATASET.txt", sep = "\t", header = TRUE, fill = TRUE)
exome.id_mappings <- read.table("/Volumes/shung/projects/gzl_targeted/comparison_to_WES/exome.id_mappings.txt", sep = "\t", header = TRUE)
gold.data <- merge(gold.data, exome.id_mappings, by = "tumor_id")

# read in calls that need to be assessed (start with frozen tumor samples) and map ids
calls.lowTC <- read.table("/Volumes/shung/projects/gzl_targeted/calls.all/calls.all.unique-union-varscan_mutect_strelka.txt", sep = "\t", header = TRUE, fill = TRUE)
target.id_mappings <- read.table("/Volumes/shung/projects/gzl_targeted/comparison_to_WES/target.id_mappings.txt", sep = "\t", header = TRUE)
calls.lowTC <- merge(calls.lowTC, target.id_mappings, by = "tumor.id")

# filter "gold data" for samples of interest (low tumor content tumors)
samples.low_tc <- read.table("/Volumes/shung/projects/gzl_targeted/roc_analysis/data/samples.low_tumor_content.txt", header = FALSE)
gold.data <- filter(gold.data, gold.data$case_id %in% samples.low_tc$V1)
calls.lowTC <- filter(calls.lowTC, calls.lowTC$case_id %in% samples.low_tc$V1)

# exclude UTR mutations
exome.UTR <- c("UTR_3_PRIME", "UTR_5_PRIME")
target.UTR <- c("5_prime_UTR_variant", "3_prime_UTR_variant")
gold.data <- filter(gold.data, !(gold.data$effect %in% exome.UTR))
calls.lowTC <- filter(calls.lowTC, !(calls.lowTC$effect %in% target.UTR))

# only look at genes in the panel
genes.panel <- read.table("/Volumes/shung/projects/gzl_exomes/data/target_panel_Agilent_genes.txt", header=FALSE)
genes.panel <- as.vector(genes.panel$V1)
gold.data <- filter(gold.data, gold.data$gene %in% genes.panel)
calls.lowTC <- filter(calls.lowTC, calls.lowTC$gene %in% genes.panel)

exome.keeps <- c("case_id", "type", "chr", "pos", "gene", "ref", "alt", "allele_freq_tumor", "var_reads_tumor", "effect")
target.keeps <- c("case_id", "type", "chr", "pos", "gene", "ref", "alt", "tumor.allele_freq", "tumor.var_reads", "effect")
gold.data <- gold.data[, exome.keeps]
calls.lowTC <- calls.lowTC[, target.keeps]

# key: tumor_id, "chr", "pos", "ref", "alt"
gold.data$key <- as.character(paste(
  gold.data$case_id,
  gold.data$chr,
  gold.data$pos,
  gold.data$ref,
  gold.data$alt,
  sep = "."
))

calls.lowTC$key <- as.character(paste(
  calls.lowTC$case_id,
  calls.lowTC$chr,
  calls.lowTC$pos,
  calls.lowTC$ref,
  calls.lowTC$alt,
  sep = "."
))

write.table(gold.data, "/Volumes/shung/projects/gzl_targeted/roc_analysis/data/gold.data.lowTC.txt", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(calls.lowTC, "/Volumes/shung/projects/gzl_targeted/roc_analysis/data/calls.lowTC.txt", sep = "\t", quote = FALSE, row.names = FALSE)

```

## Perform the ROC analysis

```{r}
# fill in values for tumor VAF and var reads into gold standard dataset

# fill in data required for ROC analysis
ROC.data <- data.frame(
  tumor_VAF.threshold_value = numeric(),
  tumor_var_reads.threshold_value = numeric(),
  TP = numeric(),
  FP = numeric(),
  FN = numeric(),
  TN = numeric(),
  exome_concordance = numeric(),
  target_concordance = numeric(),
  num_samples = numeric()
)

# Now for each threshold variable, fill in information, one threshold value at a time

# iterate through a vector of values to iterate through for tumor variant reads:
tumor_var_reads.thresholds <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15)
tumor_VAF.thresholds <- c(1, 2, 3, 3.5, 4, 4.5, 5, 5.5, 6)

# for each element in AF.normal.thresholds
for (tumor_VAF.threshold_value in tumor_VAF.thresholds) {
  for (tumor_var_reads.threshold_value in tumor_var_reads.thresholds) {
    # create a subset based on the thresholds
    calls.filtered <- subset(calls.lowTC, as.numeric(calls.lowTC$tumor.allele_freq) > tumor_VAF.threshold_value & 
                               as.numeric(calls.lowTC$tumor.var_reads) > tumor_var_reads.threshold_value) 
    calls.unfiltered <- subset(calls.lowTC, calls.lowTC$tumor.allele_freq <= tumor_VAF.threshold_value |
                                 calls.lowTC$tumor.var_reads <= tumor_var_reads.threshold_value)
    # fill in TP, TN, FP, and FN for the threshold
    # TP <- overlap of filtered calls with gold standard
    TP <- length(intersect(unique(calls.filtered$key), unique(gold.data$key)))
    # FP <- number of filtered calls that are NOT in gold standard (total filtered calls minus TP)
    FP <- nrow(calls.filtered) - TP
    # FN <- number of gold standard that are NOT in filtered calls (total gold standard minus TP)
    FN <- nrow(gold.data) - TP
    # TN <- unfiltered calls (negative dataset) not in gold (total unfiltered minus overlap of unfiltered + gold std)
    unfiltered.intersect.gold <- length(intersect(unique(calls.unfiltered$key), unique(gold.data$key)))
    TN <- nrow(calls.unfiltered) - unfiltered.intersect.gold
    # exome concordance
    exome_concordance <- TP / nrow(gold.data) * 100
    # target concordnace
    target_concordnace <- TP / nrow(calls.filtered) * 100
    # number of cases that are still retained
    num_samples <- length(unique(calls.filtered$case_id))
    
    # insert new row of data
    ROC.data <- rbind(ROC.data, data.frame(tumor_VAF.threshold_value, tumor_var_reads.threshold_value, 
                                           TP, FP, FN, TN,
                                           exome_concordance, target_concordnace, num_samples))
  }
}

# Calculate TPR and FPR
# FPR (1 - specificity) = 1 - [ TN / ( TN + FP ) ]
ROC.data$FPR <- (1 - (ROC.data$TN / (ROC.data$TN + ROC.data$FP)))*100
# TPR = TP / ( TP + FN )
ROC.data$TPR <- (ROC.data$TP / (ROC.data$TP + ROC.data$FN))*100

write.table(ROC.data, "/Volumes/shung/projects/gzl_targeted/roc_analysis/ROC_results-lowTC-tumor_VAF_and_tumor_var_reads.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

```{r}
# Now plot the ROC curve with the data (x-axis is FPR, y-axis is TPR)
#ROC.p <- ggplot(data = ROC.data, aes(x=FPR, y=TPR, color=as.factor(AF_normal.threshold_value))) + geom_point()

# with curves labelled by VAF in tumor
ROC.p <- ggplot(data = ROC.data, aes(x=FPR, y=TPR, group=as.factor(tumor_VAF.threshold_value))) +
                  geom_line(aes(color=as.factor(tumor_VAF.threshold_value))) +
                  geom_point(aes(color=as.factor(tumor_VAF.threshold_value))) +
                  labs(color = "VAF in tumor")

# delineating both factors
ROC.p2 <- ggplot(data = ROC.data, aes(x=FPR, y=TPR, group=as.factor(tumor_VAF.threshold_value))) +
                  geom_line(size=0.8, aes(color=as.factor(tumor_VAF.threshold_value))) +
                  geom_point(shape=1, aes(color=as.factor(tumor_VAF.threshold_value), size=as.factor(tumor_var_reads.threshold_value))) +
                  labs(color = "VAF in tumor", size = "Tumor variant reads")


```

```{r}
##
# Analysis of ROC curves suggests that thresholds of % AF in the normal < 0.15 and SPV < 0.175 - where are all the false positives coming from?
##
AF_normal.threshold_value <- 15
spv.threshold_value <- 0.175
# only apply the optimized thresholds (leave other parameters as minimal filters)
calls.filtered <- subset(calls.t_v_ME, as.numeric(calls.t_v_ME$allele_freq_normal) < AF_normal.threshold_value &
                           as.numeric(calls.t_v_ME$somatic_p_value) < spv.threshold_value) 

# apply optimized thresholds AND for remaining thresholds, the same values that were used previously 
calls.filtered <- subset(calls.t_v_ME, as.numeric(calls.t_v_ME$allele_freq_normal) < AF_normal.threshold_value &
                           as.numeric(calls.t_v_ME$somatic_p_value) < spv.threshold_value &
                           as.numeric(calls.t_v_ME$var_reads_tumor) > 2 & 
                           as.numeric(calls.t_v_ME$allele_freq_tumor) >= 5) 

calls.filtered$combine = as.character(paste(
  calls.filtered$tumor_id,
  calls.filtered$chr,
  calls.filtered$pos,
  calls.filtered$ref,
  calls.filtered$alt,
  sep = "."
))

#write.table(calls.filtered, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/validation/ROC_analysis/snvs.AF_norm_0.15.SPV_0.175.AF_tumor_5.var_read_3.txt", sep = "\t", quote = FALSE, row.names = FALSE)

##
# Distributions of each variable we want to test (to get an idea of what threshold values to test)
##
p.AF_normal <- ggplot(calls.t_v_ME, aes(x = allele_freq_normal, color=tumor_id)) + 
  geom_density() +
  xlim(0,10)

p.somatic_pvalue <- ggplot(calls.t_v_ME, aes(x = somatic_p_value, color=tumor_id)) + 
  geom_density()

##
```

## Application of optimized thresholds based on ROC analysis to final dataset + comparison with gold standard

```{r}
library(ggplot2)
library(RColorBrewer)

# apply optimized thresholds to T vs. ME data (p-value < 1 dataset, GMAF + effect filtered)
# this is equivalent to calls.filtered from above block of code
head(calls.filtered)

# overlap with Mutect + Strelka datasets (t vs. ME data)

# read in and process Mutect dataset
mutect.snvs <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/not_validation/mutect/mutect.all.snps.pass.txt", sep = "\t", header = TRUE)

mutect.snvs$combine = as.character(paste(
  mutect.snvs$tumor_name,
  mutect.snvs$contig,
  mutect.snvs$position,
  mutect.snvs$ref_allele,
  mutect.snvs$alt_allele,
  sep = "."
))

mutect.snvs$type = "snv"

# read in and process Strelka dataset

strelka.snvs <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/not_validation/strelka/passed.snvs.txt", sep = "\t", header = FALSE)
strelka.indels <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/not_validation/strelka/passed.indels.txt", sep = "\t", header = FALSE)

# combine snvs and indels - add mutation type as separate "type" column
strelka.indels$type = "indel"
strelka.snvs$type = "snv"
strelka.calls <- rbind(strelka.indels, strelka.snvs)

rm(strelka.indels)
rm(strelka.snvs)

# extract only columns of interest
strelka.calls <- strelka.calls[,c("V1", "V2", "V3", "V5", "V6", "type")]

# assign column names to strelka datasets
colnames(strelka.calls) <- c("sample", "chr", "pos", "ref", "alt", "type")

# remove ".markDups" from the tumor ids
strelka.calls$sample <- gsub("(.*)\\.markDups(.*)","\\1", strelka.calls$sample)

strelka.calls$combine = as.character(paste(
  strelka.calls$sample,
  strelka.calls$chr,
  strelka.calls$pos,
  strelka.calls$ref,
  strelka.calls$alt,
  sep = "."
))

# Find the intersection of varscan (snvs), mutect, and strelka (snvs).
combine.varscan.intersect.strelka <- intersect(unique(calls.filtered$combine), unique(strelka.calls$combine))
combine.intersect.all <- intersect(unique(combine.varscan.intersect.strelka), unique(mutect.snvs$combine))
snvs.intersect.all <- subset(calls.filtered, calls.filtered$combine %in% combine.intersect.all)
snvs.intersect.all$evidence <- "mutect, strelka, varscan"
#
# add indels
# Take the intersect of indels from varscan and strelka indel calls.
varscan.indels <- subset(calls.filtered, calls.filtered$type == 'indel')
strelka.indels <- subset(strelka.calls, strelka.calls$type == 'indel')
combine.varscan.intersect.strelka <- intersect(unique(varscan.indels$combine), unique(strelka.indels$combine))
indels.intersect.all <- subset(calls.filtered, calls.filtered$combine %in% combine.varscan.intersect.strelka)
indels.intersect.all$evidence <- "strelka, varscan"

# now combine snvs and indels
calls.all <- rbind(snvs.intersect.all, indels.intersect.all)

rm(calls.filtered)
rm(strelka.calls)
rm(mutect.snvs)
rm(combine.varscan.intersect.strelka)

calls.all$fun_class[calls.all$fun_class==""] <- "NONE"
calls.all$fun_class[is.na(calls.all$fun_class)] <- "NONE"

# remove the "combine" and "tumor_id" column, and rearrange columns in the process
keeps <- c("tumor_id", "type", "chr", "pos", "gene", "ref", "alt", "HGVS_DNA", "HGVS_protein",
           "allele_freq_normal", "allele_freq_tumor", "depth4_normal", "depth4_tumor", 
           "var_reads_normal", "var_reads_tumor", "effect", "impact", "fun_class", "transcript_id",
           "external_id", "somatic_p_value", "somatic_status", "gt_normal", "gt_tumor", "evidence")

calls.all <- calls.all[keeps]

# sort the calls by case, chr, then position
calls.all <- unique(arrange(calls.all, tumor_id, chr, pos))

# write the results to a text file
write.table(calls.all, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/validation/ROC_analysis/calls.T_v_ME.validation_cohort.optimized_thresholds.intersect.txt", sep = "\t", quote = FALSE, row.names = FALSE)

# read in variant calls for tumor vs. const
calls.t_v_const <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/validation/ROC_analysis/snvs.T_vs_const.validation.FINAL.txt", sep = "\t", header = TRUE, fill = TRUE)

calls.t_v_const$combine = as.character(paste(
  calls.t_v_const$tumor_id,
  calls.t_v_const$chr,
  calls.t_v_const$pos,
  calls.t_v_const$ref,
  calls.t_v_const$alt,
  sep = "."
))

# Get three datasets of interest - attached "dataset" label to each
# 1. Mutations that are found in both T vs. ME and T. vs. Blood ("Overlap")
combine.t_v_me.intersect.t_v_blood <- intersect(unique(calls.all$combine), unique(calls.t_v_const$combine))
calls.t_v_me.intersect.t_v_blood <- subset(calls.all, calls.all$combine %in% combine.t_v_me.intersect.t_v_blood)
calls.t_v_me.intersect.t_v_blood$dataset <- "Overlap"

# 2. Mutations that are found only in T vs. ME dataset ("Unique to T vs. ME")
calls.unique.t_v_me <- subset(calls.all, !(calls.all$combine %in% combine.t_v_me.intersect.t_v_blood))
calls.unique.t_v_me$dataset <- "Unique to T vs. ME"

# 3. Mutations that are found only in T vs. Blood dataset ("Unique to T vs. Blood")
calls.unique.t_v_const <- subset(calls.t_v_const, !(calls.t_v_const$combine %in% combine.t_v_me.intersect.t_v_blood))
calls.unique.t_v_const$dataset <- "Unique to T vs. Blood"

# combine above 3 datasets into one dataframe
keeps <- c("tumor_id", "type", "effect", "allele_freq_tumor", "dataset")
calls.t_v_me.intersect.t_v_blood <- calls.t_v_me.intersect.t_v_blood[keeps]
calls.unique.t_v_me <- calls.unique.t_v_me[keeps]
calls.unique.t_v_const <- calls.unique.t_v_const[keeps]
calls.combined <- rbind(calls.unique.t_v_me, calls.t_v_me.intersect.t_v_blood, calls.unique.t_v_const)

# create summary table based on tumor id + dataset type
summary.table <- as.data.frame(table(calls.combined$tumor_id, calls.combined$dataset))
colnames(summary.table) <- c("sample", "dataset", "num_mutations")
dataset.order <- c("Unique to T vs. Blood", "Overlap", "Unique to T vs. ME")
summary.table <- summary.table[order(match(summary.table$dataset, dataset.order)), ]
summary.table$dataset <- factor(summary.table$dataset, levels = summary.table$dataset)

# plot barplot, where fill (stack) of bar is defined by dataset type, each bar is a sample (e.g. GZ149T)
p <- ggplot(summary.table, aes(x=sample, y=num_mutations, fill = dataset)) +     
  geom_bar(stat = 'identity', width = 0.6) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_brewer(palette="Set1") +
  xlab("") +
  ylab("Number of mutations") 

#  geom_text(aes(label=num_mutations), vjust=1.6, color="white", size=3.5)

################################################################################
# same barplot but with original thresholds (i.e. with a total of 865 variants)
################################################################################

# load tumor vs. microenvironment
calls.t_v_me <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/validation/analysis/snvs.t_vs_ME.validation.FINAL.txt", sep = "\t", header = TRUE, fill = TRUE)

calls.t_v_me$combine = as.character(paste(
  calls.t_v_me$tumor_id,
  calls.t_v_me$chr,
  calls.t_v_me$pos,
  calls.t_v_me$ref,
  calls.t_v_me$alt,
  sep = "."
))

# extract only tumors that are in the validation cohort
tumors_validation <- as.vector(unique(calls.t_v_const$tumor_id))
calls.t_v_me <- subset(calls.t_v_me, tumor_id %in% tumors_validation)
calls.t_v_me$tumor_id <- factor(calls.t_v_me$tumor_id)

calls.all <- calls.t_v_me

# Get three datasets of interest - attached "dataset" label to each
# 1. Mutations that are found in both T vs. ME and T. vs. Blood ("Overlap")
combine.t_v_me.intersect.t_v_blood <- intersect(unique(calls.all$combine), unique(calls.t_v_const$combine))
calls.t_v_me.intersect.t_v_blood <- subset(calls.all, calls.all$combine %in% combine.t_v_me.intersect.t_v_blood)
calls.t_v_me.intersect.t_v_blood$dataset <- "Overlap"

# 2. Mutations that are found only in T vs. ME dataset ("Unique to T vs. ME")
calls.unique.t_v_me <- subset(calls.all, !(calls.all$combine %in% combine.t_v_me.intersect.t_v_blood))
calls.unique.t_v_me$dataset <- "Unique to T vs. ME"

# 3. Mutations that are found only in T vs. Blood dataset ("Unique to T vs. Blood")
calls.unique.t_v_const <- subset(calls.t_v_const, !(calls.t_v_const$combine %in% combine.t_v_me.intersect.t_v_blood))
calls.unique.t_v_const$dataset <- "Unique to T vs. Blood"

# combine above 3 datasets into one dataframe
keeps <- c("tumor_id", "type", "effect", "allele_freq_tumor", "dataset")
calls.t_v_me.intersect.t_v_blood <- calls.t_v_me.intersect.t_v_blood[keeps]
calls.unique.t_v_me <- calls.unique.t_v_me[keeps]
calls.unique.t_v_const <- calls.unique.t_v_const[keeps]
calls.combined <- rbind(calls.unique.t_v_me, calls.t_v_me.intersect.t_v_blood, calls.unique.t_v_const)

# create summary table based on tumor id + dataset type
summary.table <- as.data.frame(table(calls.combined$tumor_id, calls.combined$dataset))
colnames(summary.table) <- c("sample", "dataset", "num_mutations")
dataset.order <- c("Unique to T vs. Blood", "Overlap", "Unique to T vs. ME")
summary.table <- summary.table[order(match(summary.table$dataset, dataset.order)), ]
summary.table$dataset <- factor(summary.table$dataset, levels = summary.table$dataset)

# plot barplot, where fill (stack) of bar is defined by dataset type, each bar is a sample (e.g. GZ149T)
p <- ggplot(summary.table, aes(x=sample, y=num_mutations, fill = dataset)) +     
  geom_bar(stat = 'identity', width = 0.6) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_brewer(palette="Set1") +
  xlab("") +
  ylab("Number of mutations") 




```

